---
title: "Analysis of Coldplay Musical Trends(2000-2022)"
subtitle: "Coldplay has a postive correlation within the energy and tempo in the songs they release"
author: 
  - Abdullah Motasim
  - Elizabeth Luong
  - Yuanting Han
thanks: "Code and data are available at: [https://github.com/Luongel1/coldplay_analysis](https://github.com/Luongel1/coldplay_analysis)."
date: today
date-format: long
abstract: "This report investigates the musical evolution of British rock band Coldplay using data from Spotify. We analyze relationships between energy and tempo . Our key findings suggest a positive correlation between energy and tempo, with higher energy tracks generally having a faster tempo."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(spotifyr)
library(ggplot2)
# Load Data
coldplay <- readRDS(here::here("data/raw_data.rds"))

```


# Introduction

Music is a powerful medium that conveys emotions, moods, and energy through a combination of rhythm, melody, and tempo. In modern music, the relationship between tempo (the speed of a song) and energy (the intensity and emotional drive) is particularly significant. Different genres and musical styles use tempo and energy in varied ways to evoke specific emotional responses, making this relationship a key area of interest in understanding how music impacts listeners.

Coldplay, formed in 1997, is renowned for their broad musical range. To better understand the development of their sound over time, this report hopes to analyze how the energy and tempo on the music are correlated. 

Data for this analysis were collected from Spotify, a widely used music streaming platform that provides various audio features for each track. These features help to quantify different musical elements, allowing for an in-depth exploration of Coldplay’s discography.


The remainder of this paper is structured as follows. @sec-data discusses the tools and technologies we utilized to source and analyze our data. @sec-discussion section graphs the data  and discusses what these graphs can tell us about the bands music.



# Data {#sec-data}
All analysis was performed using R @citeR and the following packages: spotifyr @citeSpotifyr and ggplot2 @citeGgplot2. The data utilized for the analysis was sourced from the Spotify Web API @spotify_developer, which provides detailed track information for artists. The variables used in this report include:

Energy: A measure from 0.0 to 1.0 reflecting the intensity and activity of a track.
Tempo: The speed of a track in beats per minute (BPM).
*Note: @cite_spotify_audio_features used to source information on variables


```{r}
#| label: fig-valence-speech
#| fig-cap: Speechiness vs. Valence-Investigating the relationship between spoken-word elements and the emotional tone of their songs.
#| echo: false
#| warning: false
#| message: false
#| output: false

coldplay |>
  ggplot(aes(
    x = valence,
    y = speechiness,
  )) +
  geom_point(alpha = 0.5) +
  theme_minimal() +
  labs(
    x = "Valence",
    y = "Speechiness",
  )

```


# Discussion{#sec-discussion}
This section will analyze the relationship and correlations between the musics tempo and its energy

Our analysis, confirms a general positive correlation between tempo and energy in modern music. As energy levels increase, so do the tempos of the tracks. This trend is particularly clear in higher-energy songs, where fast tempos are more likely to accompany high levels of excitement and intensity. The data supports this conclusion, showing a clustering of high-energy tracks (with energy levels between 0.5 and 1.0) around higher tempo values, typically between -5 and 0 on the tempo scale.

```{r}
#| label: fig-energy-tempo
#| fig-cap: Energy vs. Tempo-Exploring how energetic and fast-paced their songs are.
#| echo: false
#| warning: false
#| message: false

coldplay |>
  ggplot(aes(
    x = energy,
    y = loudness,
  )) +
  geom_point(alpha=0.6) +
    theme_minimal() +
  labs(
    x = "Energy",
    y = "Tempo",
  )

```

## Positive Correlation Between Tempo and Energy
The scatter plot reveals that, for the majority of songs, higher energy levels are associated with faster tempos. This is especially true for songs with energy levels above 0.5, where the tempo rarely dips below -10, indicating a generally fast pace. This pattern aligns with the conventional understanding of high-energy music—genres like EDM, metal, and pop frequently use faster tempos to evoke excitement, movement, or urgency.

Data Support: The concentration of data points in the upper-right quadrant of the chart confirms this trend. These points represent high-energy tracks that align with faster tempos, reinforcing the idea that an increase in musical energy often corresponds to a more rapid rhythm.

## General Trends and Implications
In summary, the analysis confirms a strong, albeit not perfect, positive correlation between tempo and energy in modern music. High-energy songs tend to have fast tempos, while low-energy songs are more often characterized by slow tempos. However, outliers in the data suggest that tempo is not the only factor influencing the energy of a track. Musicians can manipulate non-rhythmic elements like sound design and dynamics to create high-energy music with slow tempos or low-energy music with faster tempos.
This understanding is valuable for both music producers and listeners. For producers, the data provides insight into how to balance tempo and energy when creating music to evoke specific emotional responses. For listeners, recognizing these trends helps explain why certain songs feel energetic or calming, even beyond the simple metric of tempo.




\newpage


# References


